Student Log-Data from a Randomized Evaluation of Educational Technology: A Causal Case Study

### 2020-02-10 paper.Rnw
fiiine I'll add a results table to sec 4

What would the effect be were all students to request few hints?
To answer that question, we estimate ``direct'' effects,
$\Xi=\EE[\xi(1)]=\EE[Y(1,H(0))]-\EE[Y(0,H(0))]$.
Like indirect effects, this expression includes the average potential
outcome $\EE[Y(1,H(0))]$; unlike indirect effects, it includes a term for
potential outcomes under the control condition.
Since the main contrast of $\Xi$ is due to treatment assignment,
construction of the estimator must account for the randomization
scheme.
This may be done with a regression estimator.
One option is to construct outcome $\tilde{Y}$ as, simply, $Y$ for
subjects in the control group and $\hat{Y}(1,0)$ for subjects in the
treatment group.
Then, regress it on a dummy variable for treatment group and dummy
variables for randomization block.
The clustering of subjects within schools can be accounted for in a
number of ways, such as including random intercepts for school or
computing cluster-robust standard error estimates.
Following the latter route, and using the \texttt{clubSandwich}
package in \texttt{R} \citep{clubsandwich}, we estimated a direct
effect of
\Sexpr{citopm(dir1.1ci)}.
Including covariates in the regression produces a direct effect of
\Sexpr{citopm(dir2.1ci)}.

>>>>

What would the effect be were all students to request few hints?
To answer that question, we estimate ``direct'' effects,
$\Xi=\EE[\xi(0)]=\EE[Y(1,H(0))]-\EE[Y(0,H(0))]$.
Like indirect effects, this expression includes the average potential
outcome $\EE[Y(1,H(0))]$; unlike indirect effects, it includes a term for
potential outcomes under the control condition.
Since the main contrast of $\Xi$ is due to treatment assignment,
construction of the estimator must account for the randomization
scheme.
This may be done with a regression estimator.
One option is to construct outcome $\tilde{Y}$ as, simply, $Y$ for
subjects in the control group and $\hat{Y}(1,0)$ (of Eq. \ref{eq:estY10}) for subjects in the
treatment group:
\begin{equation*}
\tilde{Y}\equiv Z\hat{Y}(1,0)+(1-Z)\hat{Y}(1,0)
\end{equation*}
Then, regress it on a dummy variable for treatment group and dummy
variables for randomization block:
\begin{equation}\label{eq:directEffReg}
\tilde{Y}_i=\alpha_{b[i]}+\hat{\Xi}Z_i+\epsilon_i
\end{equation}
where $\alpha_{b[i]}$ is a fixed intercept for subject $i$'s randomization block, $\hat{\Xi}$ is an estimate of $\Xi$ and $\epsilon_i$ is a regression error.
The clustering of subjects within schools can be accounted for in a
number of ways, such as including random intercepts for school or
computing cluster-robust standard error estimates.
Following the latter route, and using the \texttt{clubSandwich}
package in \texttt{R} \citep{clubsandwich}, we estimated a direct
effect of
\Sexpr{citopm(dir1.1ci)}.
Adding covariates to the regression \eqref{eq:directEffReg} estimates a direct effect of
\Sexpr{citopm(dir2.1ci)}.

Table \ref{tab:mediation} summarizes our mediation results.

\begin{table}
\begin{tabular}{llll}
Effect&Method&Estimate&CI\\
<<mediationTable,echo=FALSE>>=
cat(
  '$1-p_0=\\EE[H(1)-H(0)|Z=1]$ & By Definition&',round(mixProp,2),'&n/a\\\\
$TOT=\\EE[Y(Z=1,H=1)-Y(Z=1,H=0)|Z=1]$&Matching&',round(effects[['tot_eff']][['coefficients']],2),'&',printci(with(effects[['tot_eff']],c(conf.low,conf.high))),'\\\\
$\\Delta=\\EE[Y(1,H(1))-Y(1,H(0))|Z=1]$&$(1-p_0)TOT$&',round(mixProp*effects[['tot_eff']][['coefficients']],2),'&',printci(mixProp*with(effects[['tot_eff']],c(conf.low,conf.high))),'\\\\
$\\Xi=\\EE[Y(1,H(0))-Y(0,H(0))]$&Regression \\eqref{eq:directEffReg}&',round(dir1.1t[1],2),'&',printci(dir1.1ci),'\\\\
$\\Xi=\\EE[Y(1,H(0))-Y(0,H(0))]$& \\eqref{eq:directEffReg} + covariates&',round(dir2.1t[1],2),'&',printci(dir2.1ci),'\n')
@
\end{tabular}
\caption{Estimated effects in mediation analysis. }
\label{tab:mediation}
\end{table}

### 2020-02-10 paper.tex
fiiine I'll add a results table to sec 4

\documentclass{article}\usepackage[]{graphicx}\usepackage[]{color}
% maxwidth is the original width if it is less than linewidth
% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

@@

% latex table generated in R 3.6.2 by xtable 1.8-4 package
% Sun Feb  9 15:15:44 2020
\begin{table}[ht]
\centering
\begin{tabular}{rllccc}
  \hline
  &&&&\multicolumn{2}{c}{Sensitivity Intervals}\\
 Weights&Estimate&Std. Error&CI&[Pretest]&[Ethnicity] \\
 \hline
OLS & -0.18 & 0.04 & [-0.25,-0.10] & [-0.41,0.06] & [-0.30,-0.05] \\ 
  ATE & -0.12 & 0.05 & [-0.23,-0.02] & [-0.45,0.20] & [-0.29,0.04] \\ 
  TOT & -0.14 & 0.06 & [-0.25,-0.03] & [-0.47,0.19] & [-0.31,0.03] \\ 
   \hline
\end{tabular}
\caption{Estimates and sensitivity analysis of the effect of hint usage.} 
\label{tab:matchResults}
\end{table}

@@

What would the effect be were all students to request few hints?
To answer that question, we estimate ``direct'' effects,
$\Xi=\EE[\xi(1)]=\EE[Y(1,H(0))]-\EE[Y(0,H(0))]$.
Like indirect effects, this expression includes the average potential
outcome $\EE[Y(1,H(0))]$; unlike indirect effects, it includes a term for
potential outcomes under the control condition.
Since the main contrast of $\Xi$ is due to treatment assignment,
construction of the estimator must account for the randomization
scheme.
This may be done with a regression estimator.
One option is to construct outcome $\tilde{Y}$ as, simply, $Y$ for
subjects in the control group and $\hat{Y}(1,0)$ for subjects in the
treatment group.
Then, regress it on a dummy variable for treatment group and dummy
variables for randomization block.
The clustering of subjects within schools can be accounted for in a
number of ways, such as including random intercepts for school or
computing cluster-robust standard error estimates.
Following the latter route, and using the \texttt{clubSandwich}
package in \texttt{R} \citep{clubsandwich}, we estimated a direct
effect of
0.14$\pm$0.23.
Including covariates in the regression produces a direct effect of
0.18$\pm$0.19.

@@

\subsection{Stan Code}
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{verbatim}
## data{
## //Sample sizes
##  int<lower=1> N; // total number of obs (long format)
##  int<lower=1> nstud; // # students
##  int<lower=1> nsec; // # sections (problems nested in sections)
## 
## // indices
##  int<lower=1,upper=nstud> studentM[N]; //for each obs, which student?
##  int<lower=1,upper=nsec> section[N]; // for each obs, which section?
## 
## // data data
##  int<lower=0,upper=1> hint[N]; // for each obs, was a hint requested?
## 
## }
## parameters{
## 
##  vector[nstud] eta; // ability parameter
##  real<lower=0,upper=1> p1; // mixing proportion
##  //ordered[2] mu;             // locations of mixture components
##  real<upper=0> mu0; //location of mixture component 0
##  vector<lower=0>[2] sigma;  // scales of mixture components
## 
##  real delta[nsec]; // section difficulty parameter
##  //real<lower=0> sigSec; // scale of difficulty parameters
## 
## }
## transformed parameters{
##  real<lower=0> mu1; // location of mixture component 1
##  mu1= -(1-p1)/p1*mu0;
## }
## model{
##  // logs of mixing proportions to save time
##  real log_p1=log(p1);
##  real log_p0=log(1-p1);
## 
## // linear predictor for logit model
##  real linPred[N];
## 
## // priors
##  sigma ~ lognormal(0, 2);
##  mu0 ~ normal(0, 2);
## // sigSec ~ lognormal(0,2);
##  p1~ beta(2,2); // I'd rather the groups have similar sizes
## 
## // model for section difficulty parameters
##  delta~normal(0,3);//sigSec);
## 
## // Rasch model for hint requests
##  for(i in 1:N)
##   linPred[i]= eta[studentM[i]]-delta[section[i]];
## 
##  hint~bernoulli_logit(linPred);
## 
## 
## // mixture model for ability parameters
##  for(n in 1:nstud)
##   target += log_sum_exp(log_p0+normal_lpdf(eta[n]|mu0,sigma[1]),
## 			log_p1+normal_lpdf(eta[n]|mu1,sigma[2]));
## 
## }
\end{verbatim}
\end{kframe}
\end{knitrout}
%\clearpage
\section{Covariate Balance Before and After Matching}
% latex table generated in R 3.6.2 by xtable 1.8-4 package
% Sun Feb  9 15:15:45 2020
\begin{table}[ht]
\centering
\begin{tabular}{lrrrrrr}
  &\multicolumn{3}{c}{Unmatched}&\multicolumn{3}{c}{Matched}\\
 \hline
 & std.diff & z &  & std.diff & z &  \\ 
  \hline
grade9 & -0.34 & -7.05 & *** & 0.03 & 0.44 &     \\ 
  gradehigher & 0.34 & 7.05 & *** & -0.03 & -0.44 &     \\ 
  raceWhiteAsian & -0.64 & -12.83 & *** & 0.03 & 0.75 &     \\ 
  raceBlackMulti & 0.23 & 4.71 & *** & 0.01 & 0.24 &     \\ 
  raceHispAIAN & 0.51 & 10.35 & *** & -0.05 & -0.98 &     \\ 
  xirt & -0.52 & -10.54 & *** & 0.09 & 1.82 & .   \\ 
  spectypical & -0.16 & -3.32 & *** & 0.03 & 0.48 &     \\ 
  specspeced & 0.34 & 7.07 & *** & -0.10 & -1.44 &     \\ 
  specgifted & -0.14 & -3.03 & **  & 0.06 & 1.16 &     \\ 
  esl0 & -0.37 & -7.70 & *** & 0.04 & 0.63 &     \\ 
  esl1 & 0.37 & 7.70 & *** & -0.04 & -0.63 &     \\ 
  stateCT & 0.58 & 11.81 & *** & -0.00 & 0.00 &     \\ 
  stateKY & -0.36 & -7.51 & *** & -0.00 & 0.00 &     \\ 
  stateLA & -0.08 & -1.68 & .   & -0.00 & 0.00 &     \\ 
  stateMI & -0.36 & -7.51 & *** & 0.00 & 0.00 &     \\ 
  stateTX & 0.43 & 8.77 & *** & 0.00 & 0.00 &     \\ 
   \hline
\end{tabular}
\caption{Covariate balance (standardized differences and Z-scores) before and after the propensity score match.} 
\label{tab:balance}
\end{table}

>>>>

\documentclass{article}\usepackage[]{graphicx}\usepackage[]{color}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

@@

% latex table generated in R 3.4.4 by xtable 1.8-2 package
% Mon Feb 10 10:49:52 2020
\begin{table}[ht]
\centering
\begin{tabular}{rllccc}
  \hline
  &&&&\multicolumn{2}{c}{Sensitivity Intervals}\\
 Weights&Estimate&Std. Error&CI&[Pretest]&[Ethnicity] \\
 \hline
OLS & -0.18 & 0.04 & [-0.25,-0.10] & [-0.41,0.06] & [-0.30,-0.05] \\ 
  ATE & -0.12 & 0.05 & [-0.23,-0.02] & [-0.45,0.20] & [-0.29,0.04] \\ 
  TOT & -0.14 & 0.06 & [-0.25,-0.03] & [-0.47,0.19] & [-0.31,0.03] \\ 
   \hline
\end{tabular}
\caption{Estimates and sensitivity analysis of the effect of hint usage.} 
\label{tab:matchResults}
\end{table}

@@

What would the effect be were all students to request few hints?
To answer that question, we estimate ``direct'' effects,
$\Xi=\EE[\xi(0)]=\EE[Y(1,H(0))]-\EE[Y(0,H(0))]$.
Like indirect effects, this expression includes the average potential
outcome $\EE[Y(1,H(0))]$; unlike indirect effects, it includes a term for
potential outcomes under the control condition.
Since the main contrast of $\Xi$ is due to treatment assignment,
construction of the estimator must account for the randomization
scheme.
This may be done with a regression estimator.
One option is to construct outcome $\tilde{Y}$ as, simply, $Y$ for
subjects in the control group and $\hat{Y}(1,0)$ (of Eq. \ref{eq:estY10}) for subjects in the
treatment group:
\begin{equation*}
\tilde{Y}\equiv Z\hat{Y}(1,0)+(1-Z)\hat{Y}(1,0)
\end{equation*}
Then, regress it on a dummy variable for treatment group and dummy
variables for randomization block:
\begin{equation}\label{eq:directEffReg}
\tilde{Y}_i=\alpha_{b[i]}+\hat{\Xi}Z_i+\epsilon_i
\end{equation}
where $\alpha_{b[i]}$ is a fixed intercept for subject $i$'s randomization block, $\hat{\Xi}$ is an estimate of $\Xi$ and $\epsilon_i$ is a regression error.
The clustering of subjects within schools can be accounted for in a
number of ways, such as including random intercepts for school or
computing cluster-robust standard error estimates.
Following the latter route, and using the \texttt{clubSandwich}
package in \texttt{R} \citep{clubsandwich}, we estimated a direct
effect of
0.14$\pm$0.23.
Adding covariates to the regression \eqref{eq:directEffReg} estimates a direct effect of
0.18$\pm$0.19.

\begin{table}
\begin{tabular}{llll}
Effect&Method&Estimate&CI\\
$1-p_0=\EE[H(1)-H(0)|Z=1]$ & By Definition& 0.3 &n/a\\
$TOT=\EE[Y(Z=1,H=1)-Y(Z=1,H=0)|Z=1]$&Matching& -0.14 & [-0.25,-0.03] \\
$\Delta=\EE[Y(1,H(1))-Y(1,H(0))|Z=1]$&$(1-p_0)TOT$& -0.04 & [-0.07,-0.01] \\
$\Xi=\EE[Y(1,H(0))-Y(0,H(0))]$&Regression \eqref{eq:directEffReg}& 0.14 & [-0.08,0.37] \\
$\Xi=\EE[Y(1,H(0))-Y(0,H(0))]$& \eqref{eq:directEffReg} + covariates& 0.18 & [-0.01,0.38]

\end{tabular}
\caption{Estimated effects in mediation analysis. }
\label{tab:mediation}
\end{table}

@@

\subsection{Stan Code}
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{verbatim}
## data{
## //Sample sizes
##  int<lower=1> N; // total number of obs (long format)
##  int<lower=1> nstud; // # students
##  int<lower=1> nsec; // # sections (problems nested in sections)
## 
## // indices
##  int<lower=1,upper=nstud> studentM[N]; //for each obs, which student?
##  int<lower=1,upper=nsec> section[N]; // for each obs, which section?
## 
## // data data
##  int<lower=0,upper=1> hint[N]; // for each obs, was a hint requested?
## 
## }
## parameters{
## 
##  vector[nstud] eta; // ability parameter
##  real<lower=0,upper=1> p1; // mixing proportion
##  //ordered[2] mu;             // locations of mixture components
##  real<upper=0> mu0; //location of mixture component 0
##  vector<lower=0>[2] sigma;  // scales of mixture components
## 
##  real delta[nsec]; // section difficulty parameter
##  //real<lower=0> sigSec; // scale of difficulty parameters
## 
## }
## transformed parameters{
##  real<lower=0> mu1; // location of mixture component 1
##  mu1= -(1-p1)/p1*mu0;
## }
## model{
##  // logs of mixing proportions to save time
##  real log_p1=log(p1);
##  real log_p0=log(1-p1);
## 
## // linear predictor for logit model
##  real linPred[N];
## 
## // priors
##  sigma ~ lognormal(0, 2);
##  mu0 ~ normal(0, 2);
## // sigSec ~ lognormal(0,2);
##  p1~ beta(2,2); // I'd rather the groups have similar sizes
## 
## // model for section difficulty parameters
##  delta~normal(0,3);//sigSec);
## 
## // Rasch model for hint requests
##  for(i in 1:N)
##   linPred[i]= eta[studentM[i]]-delta[section[i]];
## 
##  hint~bernoulli_logit(linPred);
## 
## 
## // mixture model for ability parameters
##  for(n in 1:nstud)
##   target += log_sum_exp(log_p0+normal_lpdf(eta[n]|mu0,sigma[1]),
## 			log_p1+normal_lpdf(eta[n]|mu1,sigma[2]));
## 
## }
\end{verbatim}
\end{kframe}
\end{knitrout}
%\clearpage
\section{Covariate Balance Before and After Matching}
% latex table generated in R 3.4.4 by xtable 1.8-2 package
% Mon Feb 10 10:56:04 2020
\begin{table}[ht]
\centering
\begin{tabular}{lrrrrrr}
  &\multicolumn{3}{c}{Unmatched}&\multicolumn{3}{c}{Matched}\\
 \hline
 & std.diff & z &  & std.diff & z &  \\ 
  \hline
grade9 & -0.34 & -7.05 & *** & 0.03 & 0.44 &     \\ 
  gradehigher & 0.34 & 7.05 & *** & -0.03 & -0.44 &     \\ 
  raceWhiteAsian & -0.64 & -12.83 & *** & 0.03 & 0.75 &     \\ 
  raceBlackMulti & 0.23 & 4.71 & *** & 0.01 & 0.24 &     \\ 
  raceHispAIAN & 0.51 & 10.35 & *** & -0.05 & -0.98 &     \\ 
  xirt & -0.52 & -10.54 & *** & 0.09 & 1.82 & .   \\ 
  spectypical & -0.16 & -3.32 & *** & 0.03 & 0.48 &     \\ 
  specspeced & 0.34 & 7.07 & *** & -0.10 & -1.44 &     \\ 
  specgifted & -0.14 & -3.03 & **  & 0.06 & 1.16 &     \\ 
  esl0 & -0.37 & -7.70 & *** & 0.04 & 0.63 &     \\ 
  esl1 & 0.37 & 7.70 & *** & -0.04 & -0.63 &     \\ 
  stateCT & 0.58 & 11.81 & *** & -0.00 & 0.00 &     \\ 
  stateKY & -0.36 & -7.51 & *** & -0.00 & 0.00 &     \\ 
  stateLA & -0.08 & -1.68 & .   & -0.00 & 0.00 &     \\ 
  stateMI & -0.36 & -7.51 & *** & 0.00 & 0.00 &     \\ 
  stateTX & 0.43 & 8.77 & *** & 0.00 & 0.00 &     \\ 
   \hline
\end{tabular}
\caption{Covariate balance (standardized differences and Z-scores) before and after the propensity score match.} 
\label{tab:balance}
\end{table}
